---
title: "heatmap"
author: "lyr"
date: "2023-06-12"
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
```

## R Markdown
# Pre-process for the Heatmap use case

```{r}


data_path <- "metadata.csv"
taxonomy_path <- "taxonomy.csv"

# data: the amount of corresponding compounds for each species
data <- read.csv(data_path, check.names = F)
# taxonomy: species taxonomic data with name and taxid
taxonomy <- read.csv(taxonomy_path, check.names = F)
head(data)
# join metabolites and taxonomy data
data <- left_join(data, taxonomy, by = "taxonomy")
# get taxids to make tree by iPhylo CLI
write.csv(data$strain_taxid, "strain_taxid.csv",row.names = F, col.names = F)

## make strain_taxid the first
strain_taxid <- data$strain_taxid
# remove column "strain_taxid"
data <- data[, -which(names(data) == "strain_taxid")]
# instert column "strain_taxid" to the first column
data <- data.frame(strain_taxid, data[,1:ncol(data)], check.names = F)
colnames(data)[1] <- "taxid"
```

Then run iPhylo CLI to make tree from "strain_taxid.csv"
Note! Run it in the directory of "iphylo.py" , and modify the input and output file path to ensure that it can be found correctly
```{python}
## -f, --file: input chemicals form file
python iphylo.py phylotree -f 'strain_taxid.csv' 
```

After that, run the following code.

## Map the species-metabolite table to tree annotation

```{r}
# anno: origin CSV file output by iPhylo CLI
# iphylo_tree_items_for_anno.csv: one of the output file with the default tree name
anno <- read.csv("iphylo_tree_items_for_anno.csv")

anno$taxid <- as.numeric(anno$taxid)
data$taxid <- as.numeric(data$taxid)
anno_2 <- left_join(anno, data, by="taxid")
# remove redundant
anno_2 <- anno_2[, 1:which(names(anno_2)=="genus.y")-1]
```



```{r}
# chemical tree
nodes <- read.csv("node_info.csv", check.names = F)
all_chemical_items <- read.csv("all_chemical_items.csv", check.names = F)
node_info_joined <- left_join(nodes, all_chemical_items, by="id")
```


```{r}
chemical_tab <- readxl::read_xlsx("Supplementary_Table_1_mz-rt_library.xlsx")
compound_name_inchikey_map <- data.frame(chemical_tab$Compound, chemical_tab$inchikey)


colnames(compound_name_inchikey_map)[2] <- "inchikey"
node_info_joined <- left_join(node_info_joined, compound_name_inchikey_map, by="inchikey")
node_info_joined <- distinct(node_info_joined)

# sort
compound_order <- node_info_joined$chemical_tab.Compound

# 1-12: basic info
columns_to_order <- colnames(anno_2)[12:length(colnames(anno_2))]

# sort by columns_order
anno_2 <- anno_2[, c("id", "taxid", "domain", "kingdom.x", "phylum.x", "class.x", "order.x", "family.x", "genus.x", "species.x", "taxonomy", intersect(compound_order, columns_to_order))]

# print
# tail(colnames(anno_2), 200)

anno_1d_heatmap <- reshape2::melt(anno_2,na.rm = FALSE, measure.vars=colnames(anno_2[,12:ncol(anno_2)]),variable.name = "Metabolites",value.name = "Abundance")
anno_1d_heatmap[anno_1d_heatmap == "NA"] <- NA

# replace NA Abundance with value 0
anno_1d_heatmap$Abundance[is.na(anno_1d_heatmap$Abundance)] <- 0

# annotation_leaf output
write.csv(anno_1d_heatmap, "tree_anno_leaf.csv",row.names = F, na="NA")

```


## Column order, input in "column order" input widget in track 1
```{r}
chemical_col_order = "INDOLELACTIC ACID,ADENINE,URACIL,AGMATINE,PENTANEDECANOIC ACID,4-HYDROXYPROLINE,INDOLE-3-ETHANOL,THYMINE,N6-(DELTA2-ISOPENTENYL)-ADENINE,CITRULLINE,ORNITHINE,LACTIC ACID,2-(4-HYDROXYPHENYL)PROPIONIC ACID,CAPRYLOYL GLYCINE,2-METHYLBUTYRYLGLYCINE,LUMICHROME,DETHIOBIOTIN,SPHINGANINE,TRYPTAMINE,DEOXYCARNITINE,PHOSPHOCHOLINE,BETAINE,TRIGONELLINE,N-ACETYL-5-HYDROXYTRYPTAMINE,ADIPIC ACID,CREATINE,CREATININE,HISTAMINE,N-ACETYL-GLUTAMIC ACID,ALPHA-AMINOADIPIC ACID,1-METHYL-HYDANTOIN,THIOACETIC ACID,TAURINE,ALPHA-N-ACETYL-GLUTAMINE,N-FORMYL-METHIONINE,PHENYLETHANOLAMINE,TYRAMINE,VALINE,NEPSILON,NEPSILON,NEPSILON-TRIMETHYLLYSINE,N-ACETYL-TYROSINE,HOMOCITRULLINE,3-HYDROXY-2-METHYL-4-PYRONE,4-ACETAMIDOBUTANOIC ACID,N-METHYL-GLUTAMATE,8-HYDROXYQUINOLINE,1-METHYLGUANIDINE,HISTIDINOL,TRYPTOPHAN,N-ACETYL-ALANINE,4-AMINOBENZOIC ACID,LYSINE,OPHTHALMIC ACID,PYRIDOXAMINE,N-ACETYL-METHIONINE,N-ACETYL-LEUCINE,ISOVALERYLGLYCINE,N-ACETYL-PHENYLALANINE,ETHANOLAMINE PHOSPHATE,N-ALPHA-ACETYL-LYSINE,TRYPTOPHAN ETHYL ESTER,SPERMINE,N-ACETYLTRYPTOPHAN,ETHYL 4-AMINOBENZOIC ACID,METHIONINE,ALANINE,5-HYDROXYLYSINE,GLYCINE,PROLINE,CHOLINE,2-AMINOBENZOIC ACID,RIBOFLAVIN,PTERIN,PANTOTHENIC ACID,1-AMINOCYCLOPROPANE-1-CARBOXYLIC ACID,THREONINE,PHENYLALANINE,TYROSINE,PYROGLUTAMIC ACID,HISTIDINE,LINOLEIC ACID,ARGININE,URIDINE,CYTIDINE,GLYCYL-GLYCINE,GLUTAMINE,ASPARAGINE,ASPARTIC ACID,HYPOXANTHINE,SUCCINIC ACID,N-ACETYL-ASPARTIC ACID,N-ACETYLPUTRESCINE,CARNITINE,CYSTEINE,1,6-ANHYDRO-B-GLUCOSE,CINNAMALDEHYDE,CARNOSINE,N-ACETYLNEURAMINATE,2,4-DIHYDROXYBUTANOIC ACID,4-ETHYLOCTANOIC ACID,PHENYLACETYL-GLUTAMINE,ACETYL-CARNITINE,CAPRYLIC ACID,4-METHYLVALERIC ACID,5-HYDROXY-TRYPTOPHAN,STEARIC ACID,PYRIDOXINE,5-HYDROXYMETHYL-2-FURANCARBOXYLIC ACID,N-ACETYL-CYSTEINE,PALMITIC ACID,4-AMINOBUTANOIC ACID,4-HYDROXYBENZALDEHYDE,GLYCERIC ACID,GLUCONIC ACID,HIPPURIC ACID,SPERMIDINE,CYSTATHIONINE,O-ACETYL-SERINE,GLUTAMIC ACID,ACONITIC ACID,CYTOSINE,DIACETYL,PYRUVIC ACID,N-METHYL-ASPARTIC ACID,ASP-PHE,GLYCEROL 2-PHOSPHATE,SERINE,O-PHOSPHO-SERINE,MALIC ACID,3'-CMP,BILIVERDIN,SN-GLYCERO-3-PHOSPHOCHOLINE,GUANOSINE 3',5'-CYCLIC MONOPHOSPHATE,ADENOSINE 2',3'-CYCLIC MONOPHOSPHATE,CYTIDINE 2',3'-CYCLIC MONOPHOSPHATE,2'-DEOXYGUANOSINE 5'-MONOPHOSPHATE,METHYLMALONIC ACID,1-METHYLADENOSINE,ADENOSINE,INOSINE,GUANOSINE,S-(5'-ADENOSYL)-L-HOMOCYSTEINE,5'-METHYLTHIOADENOSINE,NICOTINAMIDE"


```

